RationalGRL: A Framework for Argumentation and Goal Modeling

### 2019-02-02 formalframework.tex

Let us explain these definitions using the example, argumentation framework, in Figure~\ref{fig:goalmodeling:arg2}, which is a subset of our RationalGRL model from Figure~\ref{fig:example-small3}, containing only arguments $A_4, A_{5},A'_{5},A_{13}$, and $A_{14}$. In this example, there are five admissible sets: $\{A'_{5}\}$, $\{A_{14}\}$, $\{A'_{5},A_{14}\}$, $\{A_4, A_{14}\}$ and $\{A_4, A'_5, A_{14}\}$. In the admissible sets that contain both $A_{4}$ and $A_{14}$, we say that $A_{14}$ defends $A_4$ against its attacker $A_{13}$\footnote{In the argumentation literature, argument $A_{14}$ is often said to \emph{reinstate} argument $A_{4}$}. Sets containing both $A_{5}$ and $A'_{5}$, or both $A_{13}$ and either $A_4$ or $A_{14}$ are not conflict free. Sets containing $A_5$ are not admissible, as they do not defend $A_5$. Similarly, sets containing $A_4$ but not $A_{14}$ are not admissible as they do not defend themselves against $A_{13}$.

>>>>

Let us explain these definitions using the example argumentation framework, in Figure~\ref{fig:goalmodeling:arg2}, which is a subset of our RationalGRL model from Figure~\ref{fig:example-small3}, containing only arguments $A_4, A_{5},A'_{5},A_{13}$, and $A_{14}$. In this example, there are five admissible sets: $\{A'_{5}\}$, $\{A_{14}\}$, $\{A'_{5},A_{14}\}$, $\{A_4, A_{14}\}$ and $\{A_4, A'_5, A_{14}\}$. In the admissible sets that contain both $A_{4}$ and $A_{14}$, we say that $A_{14}$ defends $A_4$ against its attacker $A_{13}$\footnote{In the argumentation literature, argument $A_{14}$ is often said to \emph{reinstate} argument $A_{4}$}. Sets containing both $A_{5}$ and $A'_{5}$, or both $A_{13}$ and either $A_4$ or $A_{14}$ are not conflict free. Sets containing $A_5$ are not admissible, as they do not defend $A_5$. Similarly, sets containing $A_4$ but not $A_{14}$ are not admissible as they do not defend themselves against $A_{13}$.

### 2019-02-02 overview.tex

In the rest of this section we discuss the individual parts of the GRL framework. In Section~\ref{sect:overview:as}, we continue our discussion of the argument schemes and critical questions for practical reasoning and argumentation, fitting these schemes and questions into our framework. In Section~\ref{sect:overview:lang}, we discuss the language for RationalGRL models and we provide a metamodel, linking the new concepts of our language to GRL. In Section~\ref{sect:overview:examples}, we provide extensive examples from our case study.

@@

\begin{figure*}[t]
\includegraphics[width=\textwidth]{metamodel/metamodel2}
\caption{The RationalGRL metamodel. The \emph{Goal model construction} package (bottom) is the GRL metamodel. The \emph{Practical reasoning \& argumentation} package is the RationalGRL extension}
\label{fig:metamodel}
\end{figure*}

@@

The practical reasoning and argumentation package depicts the concepts we introduced in Section~\ref{sect:overview:as}. An \textsf{ArgumentScheme} represents a scheme containing variables. When an argument scheme is instantiated, we obtain an \textsf{Argument}. Therefore, each argument is associated with at most one scheme, but a scheme can be instantiated in multiple ways. A \textsf{RationalGRLmodel} is composed out of arguments and attack relations.

The \textsf{OperationTypes} in the Argumentation package correspond to the operations we informally introduced in Section~\ref{sect:overview}. These operations are performed by instantiating an argument scheme or answering a critical question in a certain way. An \textsf{INTRO} operation introduces a new RationalGRL element. A \textsf{DISABLE} operation creates a new argument that attacks another argument or GRL element, effectively disabling it. The \textsf{REPLACE} operation replaces a RationalGRL element with a new element. Instantiating an argument scheme from Table~\ref{table:argument-schemes} always leads to an \textsf{INTRO} operation, that is, it always introduces a new RationalGRL element. Answering a critical question can have different effects depending on the critical question and the answer. Table~\ref{table:argument-schemes} shows these effects for the different critical questions and answers. For example, answering CQ0 with ``no'' disables the argument based on AS0.

There are two important links between the \emph{practical reasoning \& argumentation} and \emph{goal model construction} packages. First, each \textsf{GRLModelElement} is an \textsf{Argument}. This means that each model element inherits the \textsf{AcceptStatus} as well, allowing GRL elements to be accepted or rejected. This, furthermore, means that argument schemes can be applied to all GRL elements, capturing the intuition that each GRL element can be regarded as an instantiated argument scheme. Note that besides arguments about elements of the GRL model, we also have a \textsf{GenericArgument} which is simply a counter-argument to an existing argument that does not relate to any of the GRL elements. Finally, the relation between \textsf{GRLModelElement} and \textsf{Argument} means that, as we already briefly indicated when discussing the framework in Figure~\ref{fig:rationalgrl-framework}, the class of \textsf{RationalGRLmodel} is a superclass of \textsf{GRLmodel}: besides arguments about GRL elements, we can also have arguments that does not relate to any of the GRL elements.

@@

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{img/fig_example_AS.pdf}
\caption{RationalGRL model (right) with instantiated argument schemes (left): Introducing new elements (operation \textsf{INTRO)}}. 
\label{fig:example_AS}
\end{figure}

@@

\begin{figure}[b]
\centering
\includegraphics[]{img/fig_example_disable.pdf}
\caption{Disabling elements (operation \textsf{DISABLE)}.}
\label{fig:example_disable}
\end{figure}

The excerpt from transcript $t_1$ used in this example is shown in Table~\ref{table:transcripts:traffic-light} in Appendix~\ref{sect:transcripts:excerpts}. In this example, participants first sum up functionality of the traffic simulator, which can be captured as instantiations of AS2. On left side of Figure~\ref{fig:example_disable} one such instantiation is shown, which leads to the addition of the task \emph{Add traffic light} in the RationalGRL model on the right side of Figure~\ref{fig:example_disable}. However, participant P1 notes that the problem description states that all intersections have traffic lights by default, so the task \emph{Add traffic light} is not necessary. This is captured using critical question CQ11. A negative answer to this question (cf. Table~\ref{table:argument-schemes}) should disable the original argument based on AS2 by attacking it. On the left side of Figure~\ref{fig:example_disable} a new argument (CQ11) attacks the original argument based on argument scheme AS2. This new argument is also added to the RationalGRL model on the right of Figure~\ref{fig:example_disable}, where it attacks the original task \emph{Add traffic light}. This attack leads to the original argument being \emph{rejected} (cf. Section~\ref{sect:background:pras} and Section~\ref{sect:formalframework:rationalgrl}), indicated by being greyed out. As a result of this, the corresponding GRL task \emph{Add traffic light} is also disabled.

@@

\begin{figure}[t]
\centering
\includegraphics[]{img/fig_example_replace.pdf}
\caption{Replacing an element (operation \textsf{REPLACE)}.}
\label{fig:examples:replace}
\end{figure}

@@

\begin{figure}[b]
\centering
\includegraphics[]{img/fig_example_rename.pdf}
\caption{Renaming an element name (operation \textsf{REPLACE)}.}
\label{fig:examples:clarification}
\end{figure}

@@

\begin{figure}[t]
\centering
\includegraphics[]{img/reinstate1.pdf}
\caption{Disabling an element (operation \textsf{DISABLE)}.}
\label{fig:examples:relevant-actor}
\end{figure}

The excerpt from transcript $t_3$ used in this example is shown in Table~\ref{table:transcript:irrelevant-actor} in Appendix~\ref{sect:transcripts:excerpts}. Actor \emph{Development Team} is introduced with an argument based on AS0 (Figure~\ref{fig:examples:relevant-actor}). This is then attacked by arguing that the professor will develop the software, so there will not be any development team (CQ0).

\begin{figure}[b]
\centering
\includegraphics[]{img/reinstate2.pdf}
\caption{Defending an element by attacking its disabling attacker (operation \textsf{DISABLE)}.}
\label{fig:examples:relevant-actor2}
\end{figure}

>>>>

In the rest of this section we discuss the individual parts of the RationalGRL framework. In Section~\ref{sect:overview:as}, we continue our discussion of the argument schemes and critical questions for practical reasoning and argumentation, fitting these schemes and questions into our framework. In Section~\ref{sect:overview:lang}, we discuss the language for RationalGRL models and we provide a metamodel, linking the new concepts of our language to GRL. In Section~\ref{sect:overview:examples}, we provide extensive examples from our case study.

@@

\begin{figure*}[t]
\includegraphics[width=\textwidth]{metamodel/metamodel2}
\caption{The RationalGRL Metamodel. The \emph{Goal model construction} package (bottom) is the GRL metamodel. The \emph{Practical reasoning \& argumentation} package is the RationalGRL extension}
\label{fig:metamodel}
\end{figure*}

@@

The practical reasoning and argumentation package depicts the concepts we introduced in Section~\ref{sect:overview:as}. An \textsf{ArgumentScheme} represents a scheme containing variables. When an argument scheme is instantiated, we obtain an \textsf{Argument}. Therefore, each argument is associated with at most one scheme, but a scheme can be instantiated in multiple ways. A \textsf{RationalGRLmodel} is composed of arguments and attack relations.

The \textsf{OperationTypes} in the argumentation package correspond to the operations we informally introduced in Section~\ref{sect:overview}. These operations are performed by instantiating an argument scheme or answering a critical question in a certain way. An \textsf{INTRO} operation introduces a new RationalGRL element. A \textsf{DISABLE} operation creates a new argument that attacks another argument or GRL element, effectively disabling it. The \textsf{REPLACE} operation replaces a RationalGRL element with a new element. Instantiating an argument scheme from Table~\ref{table:argument-schemes} always leads to an \textsf{INTRO} operation, that is, it always introduces a new RationalGRL element. Answering a critical question can have different effects depending on the critical question and the answer. Table~\ref{table:argument-schemes} shows these effects for the different critical questions and answers. For example, answering CQ0 with ``no'' disables the argument based on AS0.

There are two important links between the \emph{practical reasoning \& argumentation} and \emph{goal model construction} packages. First, each \textsf{GRLModelElement} is an \textsf{Argument}. This means that each model element inherits the \textsf{AcceptStatus} as well, allowing GRL elements to be accepted or rejected. This, furthermore, means that argument schemes can be applied to all GRL elements, capturing the intuition that each GRL element can be regarded as an instantiated argument scheme. Note that besides arguments about elements of the GRL model, we also have a \textsf{GenericArgument} which is simply a counter-argument to an existing argument that does not relate to any of the GRL elements. Finally, the relation between \textsf{GRLModelElement} and \textsf{Argument} means that, as we already briefly indicated when discussing the framework in Figure~\ref{fig:rationalgrl-framework}, the class of \textsf{RationalGRLmodel} is a superclass of \textsf{GRLmodel}: besides arguments about GRL elements, we can also have arguments that do not relate to any of the GRL elements.

@@

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{img/fig_example_AS.pdf}
\caption{RationalGRL Model (right) with Instantiated Argument Schemes (left): Introducing New Elements (Operation \textsf{INTRO)}}. 
\label{fig:example_AS}
\end{figure}

@@

\begin{figure}[b]
\centering
\includegraphics[]{img/fig_example_disable.pdf}
\caption{Disabling Elements (Operation \textsf{DISABLE)}.}
\label{fig:example_disable}
\end{figure}

The excerpt from transcript $t_1$ used in this example is shown in Table~\ref{table:transcripts:traffic-light} in Appendix~\ref{sect:transcripts:excerpts}. In this example, participants first sum up functionality of the traffic simulator, which can be captured as instantiations of AS2. On left side of Figure~\ref{fig:example_disable} one such instantiation is shown, which leads to the addition of the task \emph{Add traffic light} in the RationalGRL model on the right side of Figure~\ref{fig:example_disable}. However, participant P1 notes that the problem description states that all intersections have traffic lights by default, so the task \emph{Add traffic light} is not necessary. This is captured using critical question CQ11. A negative answer to this question (cf. Table~\ref{table:argument-schemes}) should disable the original argument based on AS2 by attacking it. On the left side of Figure~\ref{fig:example_disable} a new argument (CQ11) attacks the original argument based on argument scheme AS2. This new argument is also added to the RationalGRL model on the right-hand side of Figure~\ref{fig:example_disable}, where it attacks the original task \emph{Add traffic light}. This attack leads to the original argument being \emph{rejected} (cf. Section~\ref{sect:background:pras} and Section~\ref{sect:formalframework:rationalgrl}), indicated by being greyed out. As a result of this, the corresponding GRL task \emph{Add traffic light} is also disabled.

@@

\begin{figure}[t]
\centering
\includegraphics[]{img/fig_example_replace.pdf}
\caption{Replacing an Element (Operation \textsf{REPLACE)}.}
\label{fig:examples:replace}
\end{figure}

@@

\begin{figure}[b]
\centering
\includegraphics[]{img/fig_example_rename.pdf}
\caption{Renaming an Element Name (Operation \textsf{REPLACE)}.}
\label{fig:examples:clarification}
\end{figure}

@@

\begin{figure}[t]
\centering
\includegraphics[]{img/reinstate1.pdf}
\caption{Disabling an Element (Operation \textsf{DISABLE)}.}
\label{fig:examples:relevant-actor}
\end{figure}

The excerpt from transcript $t_3$ used in this example is shown in Table~\ref{table:transcript:irrelevant-actor} in Appendix~\ref{sect:transcripts:excerpts}. Actor \emph{Development Team} is introduced with an argument based on AS0 (Figure~\ref{fig:examples:relevant-actor}). This is then attacked by arguing that the professor will develop the software, thus, there will not be any development team (CQ0).

\begin{figure}[b]
\centering
\includegraphics[]{img/reinstate2.pdf}
\caption{Defending an Element by Attacking its Disabling Attacker (Operation \textsf{DISABLE)}.}
\label{fig:examples:relevant-actor2}
\end{figure}

### 2019-02-02 tool.tex

We propose the development process shown in Figure~\ref{fig:rationalgrl-methodology} to develop a RationalGRL model, a version of which was presented at the 2017 iStar workshop \cite{ghanavatiMethodology}. Here we assume that the initial GRL models have been created based on the requirements specification documents and the discussions of the stakeholders. The rest of the steps are as follows:

@@

The final requirement of our framework is that it has tool support (\textbf{requirement 5}). This is important for various reasons: i) although there are various approaches attempting to combine goal modeling with argumentation, we are not aware of existing tool support (see Section~\ref{sect:discussion}), ii) it allows us to do user tests in future research, exploring the difference between GRL and RationalGRL, iii) tool support is an excellent demonstration to show that our formal framework and algorithms actually work. In this section we briefly highlight some features of the tool, and we explain some of its current limitations. The tool can be accessed from:

@@

GRL is implemented in the open-source tool jUCMNav~\cite{jUCMNav}, which is actively maintained\footnote{\url{http://jucmnav.softwareengineering.ca/foswiki/ProjetSEG}}. jUCMNav currently has over 2,000 commits made by over 40 contributors, representing over 200,000 lines of code. jUCMNav is used actively in research, and the tool has been extended in many directions over the past few years, such as for modeling and analyzing security misuse cases~\cite{daramola2012ontology}, supporting activity theory \cite{georg2015synergy}, combined goal and feature model reasoning~\cite{liu2014combined}, and enterprise architecture modeling \cite{marosin-etal:caise2016}. Furthermore, jUCMNav includes Use Case modeling and contains various GRL evaluation algorithms with which the satisfiability of goals can be determined.

The jUCMNav tool is implemented as an Eclipse plugin. While we aim to integrate our framework into jUCMNav in the future, a web-based version of RationalGRL (and GRL) is more suited for current purposes. Since RationalGRL does not make use of the many features of jUCMNav, we believe a light-weight version is sufficient for our current aims, and may benefit the community in general. Our tool is thus not meant to replace jUCMNav. For simplicity we have a web-based version, and it has the functionality to export to jUCMNav.

@@

In total there are four different types of details panes, which we now explain in turn.

@@

\paragraph{Critical question details pane} The details panel can be used to answer the critical questions of the RationalGRL framework (Table~\ref{table:argument-schemes}): Figure~\ref{fig:tool:overview}, for example, shows two critical questions for the goal IE ``Generate cars'', namely ``Can the goal be realized?'' (CQ3) and ``Is the goal relevant/useful?'' (CQ11). As an example, consider Figure~\ref{fig:tool:cqdetails}, where the softgoal ``Easy to use'' is questioned with the relevancy question (CQ11). It is possible to select the answer and provide an explanation for the answer.

Assume that in the example of Figure~\ref{fig:tool:cqdetails} the user selects ``No'' and clicks the ``Answer question'' button. A new argument is then automatically created that attacks the softgoal and the details pane shows the critical question as answered (Figure~\ref{fig:tool:cqeffect}). This is an implementation of a \textsf{DISABLE} algorithm (Section~\ref{sect:formalframework:disable}) similar to Algorithm~\ref{alg:actor-not-relevant}: a new argument is added that attacks the existing argument.

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.8\columnwidth]{img/tool/details_softgoal}
\caption{Critical question details pane for ``Is the softgoal relevant/useful?''}
\label{fig:tool:cqdetails}
\end{figure}

\begin{figure}[b]
\centering
\includegraphics[width=\columnwidth]{img/tool/attack_softgoal}
\caption{Effect of answering CQ11: ``Is the softgoal relevant/useful?'' with ``No''}
\label{fig:tool:cqeffect}
\end{figure}

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.8\columnwidth]{img/tool/reinstate_softgoal}
\caption{Attacking an argument with a generic argument ``Necessary''}
\label{fig:tool:argument}
\end{figure}

@@

Our translation and export function uses the argument acceptability as a way of determining the GRL model. Take for example, the RationalGRL model in Figure~\ref{fig:tool:multipleattack}. A positive contribution is added from ``Keep same cars'' to ``Simple design''. More inportantly, an argument ``not enough CPU'' has been added to the model. This argument attacks the ``Realistic simulation'' softgoal and the ``Create new cars'' task, arguing that there is not enough processing power for either of these to be feasible in the traffic simulator. Now, if we export this to GRL, the GRL elements that are rejected or disabled (greyed out in Figure~\ref{fig:tool:multipleattack}) are not included. This can be seen in Figure~\ref{fig:tool:multipleattack1}, where the jUCMNav GRL model that was exported from the RationalGRL tool is shown. The pair of figures \ref{fig:tool:multipleattack} and \ref{fig:tool:multipleattack1} also nicely shows the added value of RationalGRL: Figure~\ref{fig:tool:multipleattack} shows that there can be a larger discussion and rationalization underlying even a fairly simple goal model such as the one in Figure~\ref{fig:tool:multipleattack1}.

>>>>

We propose the development process shown in Figure~\ref{fig:rationalgrl-methodology} to develop a RationalGRL model, a version of which was presented at the 2017 iStar workshop \cite{ghanavatiMethodology}. Here, we assume that the initial GRL models have been created based on the requirements specification documents and the discussions of the stakeholders. The rest of the steps are as follows:

@@

The final requirement of our framework is that it has tool support (\textbf{requirement 5}). This is important for various reasons: i) although there are various approaches attempting to combine goal modeling with argumentation, we are not aware of an existing tool support (see Section~\ref{sect:discussion}), ii) it allows us to conduct user studies in the next section, exploring the usability and usefulness of the RationalGRL framework compared to other goal modeling languages, and iii) tool support is an excellent demonstration to show that our formal framework and algorithms actually work. In this section, we briefly highlight some features of the tool, and we explain some of its current limitations. The tool can be accessed from:

@@

GRL is implemented in the open-source tool jUCMNav~\cite{jUCMNav}, which is actively maintained\footnote{\url{http://jucmnav.softwareengineering.ca/foswiki/ProjetSEG}}. jUCMNav currently has over 2,000 commits made by over 40 contributors, representing over 200,000 lines of code. jUCMNav is used actively in research, and the tool has been extended in many directions over the past few years, such as for modeling and analyzing security misuse cases~\cite{daramola2012ontology}, supporting activity theory \cite{georg2015synergy}, combined goal and feature model reasoning~\cite{liu2014combined}, and enterprise architecture modeling \cite{marosin-etal:caise2016}. Furthermore, jUCMNav includes Use Case Maps for modeling business processes and scenarios and contains various GRL evaluation algorithms with which the satisfiability of goals can be determined.

The jUCMNav tool is implemented as an Eclipse plugin. While we aim to integrate our framework into jUCMNav in the future, a web-based version of RationalGRL (and GRL) is more suited for current purposes. Since RationalGRL does not make use of the many features of jUCMNav, we believe a light-weight version is sufficient for our current aims, and may benefit the community in general. Our tool is thus not meant to replace jUCMNav. For simplicity, we have a web-based version, and it has the functionality to export to jUCMNav.

@@

In total, there are four different types of details panes, which we now explain in turn.

@@

\paragraph{Critical question details pane} The details pane can be used to answer the critical questions of the RationalGRL framework (Table~\ref{table:argument-schemes}): Figure~\ref{fig:tool:overview}, for example, shows two critical questions for the goal IE ``Generate cars'', namely ``Can the goal be realized?'' (CQ3) and ``Is the goal relevant/useful?'' (CQ11). As an example, consider Figure~\ref{fig:tool:cqdetails}, where the softgoal ``Easy to use'' is questioned with the relevancy question (CQ11). It is possible to select the answer and provide an explanation for the answer.

Assume that in the example of Figure~\ref{fig:tool:cqdetails}, the user selects ``No'' and clicks the ``Answer question'' button. A new argument is then automatically created that attacks the softgoal and the details pane shows the critical question as answered (Figure~\ref{fig:tool:cqeffect}). This is an implementation of a \textsf{DISABLE} algorithm (Section~\ref{sect:formalframework:disable}) similar to Algorithm~\ref{alg:actor-not-relevant}: a new argument is added that attacks the existing argument.

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.8\columnwidth]{img/tool/details_softgoal}
\caption{Critical Question Details Pane for ``Is the softgoal relevant/useful?''}
\label{fig:tool:cqdetails}
\end{figure}

\begin{figure}[b]
\centering
\includegraphics[width=\columnwidth]{img/tool/attack_softgoal}
\caption{Effect of Answering CQ11: ``Is the softgoal relevant/useful?'' with ``No''}
\label{fig:tool:cqeffect}
\end{figure}

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.8\columnwidth]{img/tool/reinstate_softgoal}
\caption{Attacking an Argument with a Generic Argument ``Necessary''}
\label{fig:tool:argument}
\end{figure}

@@

Our translation and export function uses the argument acceptability as a way of determining the GRL model. Take for example, the RationalGRL model in Figure~\ref{fig:tool:multipleattack}. A positive contribution is added from ``Keep same cars'' to ``Simple design''. More importantly, an argument ``not enough CPU'' has been added to the model. This argument attacks the ``Realistic simulation'' softgoal and the ``Create new cars'' task, arguing that there is not enough processing power for either of these to be feasible in the traffic simulator. Now, if we export this to GRL, the GRL elements that are rejected or disabled (greyed out in Figure~\ref{fig:tool:multipleattack}) are not included. This can be seen in Figure~\ref{fig:tool:multipleattack1}, where the jUCMNav GRL model that was exported from the RationalGRL tool is shown. The pair of figures \ref{fig:tool:multipleattack} and \ref{fig:tool:multipleattack1} also nicely shows the added value of RationalGRL: Figure~\ref{fig:tool:multipleattack} shows that there can be a larger discussion and rationalization underlying even a fairly simple goal model such as the one in Figure~\ref{fig:tool:multipleattack1}.

### 2019-02-02 validation.tex

\section{RationalGRL evaluation experiment}
\label{sect:validation}

@@

\begin{itemize}
\item[H1] The additions of RationalGRL -- arguments, critical questions and determination of the status of arguments -- are a useful addition to standard goal modelling in GRL.
\item[H2] The additions of RationalGRL make it easier to express, determine the effect of and communicate to other stakeholders one's opinions and beliefs about a goal model.
\end{itemize}

@@

The idea was to have our participants perform a small modelling task, and then ask them what they thought of RationalGRL. Thus, our experiment consisted of three parts:

@@

For part 3, we provided the participants with a survey\footnote{https://goo.gl/forms/fDSMUnAV20wy7kbY2}. The survey starts with general questions about the participant (years of experience etc.). The second part of the survey concerns the specific features of RationalGRL, asking participants to rate RationalGRL on a Likert scale from 1 (very useless) to 5 (very useful):
\begin{itemize}
\item[Q1] Do you think the arguments and counterarguments of RationalGRL are a useful extension to standard goal modeling?
\item[Q2] Do you think the critical questions and answers in the details pane are a useful extension to standard goal modeling?
\item[Q3] Do you think the automatic determination of the status of arguments and elements is a useful extension to standard goal modeling?
\end{itemize}
The following questions ask the participants to rate RationalGRL vs. standard goal modelling on a Likert scale from 1 (much more difficult) to 5 (much easier):
\begin{itemize}
\item[Q4] Do you think using RationalGRL instead of a standard goal modeling language makes it easier or more difficult to for someone to express beliefs and opinions in a goal model?
\item[Q5] Do you think using RationalGRL instead of a standard goal modeling language makes it easier or more difficult to for someone to determine the effect of beliefs and opinions on the resulting goal model?
\item [Q6] Do you think using RationalGRL instead of a standard goal modeling language makes it easier or more difficult to for someone who is not the original author to understand the goal model? 
\end{itemize}
We also ask two open questions regarding the strengths and weaknesses of RationalGRL compared to GRL.

The survey also contains questions about the RationalGRL tool user experience, such as whether it was easy or difficult to get started with the tool and to model the example case. We also ask open questions about the strengths and weaknesses of the tool, possible improvements and whether the participant thinks the tool could in the future be used in practice.

@@

\subsection{Analysis or results}\label{sect:validation:analysis}

>>>>

\section{RationalGRL Evaluation Experiment}
\label{sect:validation}

@@

\begin{itemize}
\item[H1] The additions of RationalGRL -- arguments, critical questions and determination of the status of arguments -- are a useful addition to standard goal modeling in GRL.
\item[H2] The additions of RationalGRL make it easier to express, determine the effect of and communicate to other stakeholders one's opinions and beliefs about a goal model.
\end{itemize}

@@

The idea was to have our participants perform a small modeling task, and then ask them what they thought of RationalGRL. Thus, our experiment consisted of three parts:

@@

For part 3, we provided the participants with a survey\footnote{https://goo.gl/forms/fDSMUnAV20wy7kbY2}. The survey starts with general questions about the participant (years of experience etc.). The second part of the survey concerns the specific features of RationalGRL, asking participants to rate RationalGRL on a Likert scale from 1 (very useless) to 5 (very useful):
\begin{itemize}
\item[Q1] Do you think the arguments and counterarguments of RationalGRL are a useful extension to standard goal modeling?
\item[Q2] Do you think the critical questions and answers in the details pane are a useful extension to standard goal modeling?
\item[Q3] Do you think the automatic determination of the status of arguments and elements is a useful extension to standard goal modeling?
\end{itemize}
The following questions ask the participants to rate RationalGRL vs. standard goal modeling language  on a Likert scale from 1 (much more difficult) to 5 (much easier):
\begin{itemize}
\item[Q4] Do you think using RationalGRL instead of a standard goal modeling language makes it easier or more difficult to for someone to express beliefs and opinions in a goal model?
\item[Q5] Do you think using RationalGRL instead of a standard goal modeling language makes it easier or more difficult to for someone to determine the effect of beliefs and opinions on the resulting goal model?
\item [Q6] Do you think using RationalGRL instead of a standard goal modeling language makes it easier or more difficult to for someone who is not the original author to understand the goal model? 
\end{itemize}
We also ask two open questions regarding the strengths and weaknesses of RationalGRL compared to GRL. The survey further contains questions about the RationalGRL tool user experience, such as whether it was easy or difficult to get started with the tool and to model the example case. We also ask open questions about the strengths and weaknesses of the tool, possible improvements and whether the participant thinks the tool could in the future be used in practice.

@@

\subsection{Analysis of Results}\label{sect:validation:analysis}
